Compositional Quantitative Resource Analysis of C Programs

### 2014-06-25 paper/main.tex
more layout
ramblings on logical variables

\pref{fig:logic} shows the inference rules of the quantitative logic.
The rules are slightly simplified in comparison to the implemented
rules in Coq.  The main difference is that the presented version do
not formalize an auxiliary state that is needed to relate pre and
postconditions (see the following paragraph for more details).

@@

In the rule {\sc L:Assert}, we use the notation
$\code{istrue}~\sem{e}_\state \implies Q + M_a$ to express that we
require potential $Q + M_a$ in the precondition if $e$ evaluates to
$\mathit{true}$ in the current program state.  If $e$ evaluates to
$\mathit{false}$ then the potential in the precondition can be
arbitrary since the program is terminated.

In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
be arbitrary since it is unreachable.  Instead, we have to justify the
potential functions $B$ and $R$ that hold after a \code{return} and a
\code{break}, respectively.  In {\sc L:Break}, we require to have
potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
cost of \code{break} and $B$ to pay for the potential after the
\code{break}.  In ${\sc L:Return}$ we only require to have potential
$R$ in the precondition to pay for the potential after the
\code{return}.  The reason is that we found it to be more convenient
to account for the execution cost $M_r$ of the return in rule {\sc
  L:Call} for function calls.

@@

There are two weakening rules available in the quantitative Hoare
logic.  The framing rule {\sc L:Frame} is designed to weaken a
statement by stating that if $S$ needs $P$ bytes to run and leaves $Q$
bytes free at its end, then it can very well run with $P + c$ bytes
and return $Q + c$ bytes. It comes very handy when we want to prove
tight bounds using the $\max$ function as demonstrated
in~\pref{fig:example_derivation}. The consequence {\sc L:Conseq} rule
is directly imported from classical Hoare logics except that instead
of using the logical implication $\Rightarrow$ we use the quantitative
$\ge$. This rule indeed weakens the statement since it requires more
resource to run the statement and yields less than what has been
proved to be available after its termination.

\paragraph{Auxiliary State and Shallow Embedding}

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule[leftskip=1.5cm]{Q:IncP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
\\ M = M_u + M_e(x {+} y)
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[rightskip=1.5cm]{Q:IncM}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
\\ M = M_u + M_e(x {-} y)
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Inc}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
\\\\ M = M_u + M_e(x {\pm} y)
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \mathbb Q_0^+
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
%
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\and \Rule{Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}

>>>>

\pref{fig:logic} shows the inference rules of the quantitative logic.
The rules are slightly simplified in comparison to the implemented
rules in Coq.  The main difference is that the presented version does
not formalize the heap operations.

@@

In the rule {\sc L:Assert}, we use the notation
$\code{istrue}~\sem{e}_\state \implies Q + M_a$ to express that we
require potential $Q + M_a$ in the precondition if $e$ evaluates to
$\mathit{true}$ in the current program state.  If $e$ evaluates to
$\mathit{false}$ then the potential in the precondition can be
arbitrary since the program will be terminated.

In the rules {\sc L:Break} and {\sc L:Return}, the postcondition can
be arbitrary since it is unreachable.  Instead, we have to justify the
potential functions $B$ and $R$ that hold after a \code{return} and a
\code{break}, respectively.  In {\sc L:Break}, we require to have
potential $M_b+B$ in the precondition: $M_b$ to pay for the execution
cost of \code{break} and $B$ to pay for the potential after the
\code{break}.  In {\sc L:Return} we only require to have potential
$R$ in the precondition to pay for the potential after the
\code{return}.  The reason is that we found it to be more convenient
to account for the execution cost $M_r$ of the return in rule {\sc
  L:Call} for function calls.

@@

There are two weakening rules available in the quantitative Hoare
logic.  The framing rule {\sc L:Frame} is designed to weaken a
statement by stating that if $S$ needs $P$ bytes to run and leaves $Q$
bytes free at its end, then it can very well run with $P + c$ bytes
and return $Q + c$ bytes. It comes very handy when we want to prove
tight bounds using the $\max$ function as demonstrated
in~\pref{fig:example_derivation}. The consequence {\sc L:Conseq} rule
is directly imported from classical Hoare logics except that instead
of using the logical implication $\implies$ we use the quantitative
$\ge$. This rule indeed weakens the statement since it requires more
resource to run the statement and yields less than what has been
proved to be available after its termination.

\paragraph{Logical Variables and Shallow Embedding}

It is often necessary, when specifying a program, to relate certain
invariants in the pre- and postcondition.  A standard solution of
Hoare logics to this problem is \emph{logical variables} (reference XXX).  These
additional variables also mentioned in the literature as auxiliary
state are constant across the derivation.  For example, if we note
$Z$ such a logical variable, we can specify the function
$\code{double}()$ which doubles the global variable $x$ as
$$
  \htriple{z = Z}{\code{double}()}{x = 2 \cdot Z}.
$$

When formalizing Hoare logics in a proof assistant one can either
fully specify the syntax and semantics of assertions and hence get a
deep embedding, or use the assertions of the host theory to get a
shallow embedding.  For simplicity, we used the latter approach in our
development.  This choice makes it possible to have logical variables
almost for free: we can simply use the variable and binding mechanisms
of Coq, our host theory.  When an logical variable is needed we
introduce it using a universal quantifier of Coq before deriving triples
in the logic, so the Coq theorem for the above example would look like
%
\begin{lstlisting}
Theorem dbl_triple: forall Z,
  htriple ($\lambda\sigma$.$\sigma$(x) = Z) ($\code{double}()$) ($\lambda\sigma$.$\sigma$(x) = 2*Z).
\end{lstlisting}

However, this trick alone is often not sufficient when working with a
recursive function $f$.  In that case we apply the {\sc L:Extend} rule
of the logic to assume the specification of $f$ then proceed to prove
the function body with this induction hypothesis.  But in this process
it can be the case that the induction hypothesis needs to be used with
a different value of a logical variable.  To cope with this problem,
assumptions in the function context $\Delta$ are universally quantified
over logical variables and the {\sc L:Extend} rule once again leverages
the host proof assistant to require that the triple on $f$'s body is
proved for every possible logical variable $y$.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
%
\\ \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule{Q:IncP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:IncM}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Inc}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
\\\\ M = M_u + M_e(x {\pm} y)
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \mathbb Q_0^+
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\and \Rule{Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}
